---
title: "projectRoads"
author: "Modified and parially copied from Kyle Lochhead & Tyler Muhly's 2018 CLUS example"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette provides a tutorial on the use of the `roads` package for the spatial simulation of future roads development under a given resource development scenario.  This tutorial, which borrows heavily from a demonstration written by Kyle Lochhead and Tyler Muhly in 2018, will focus primarily on the `projectRoads` method of the `roads` package, as it is responsible for performing the simulation. Example datasets from Kyle Lochhead and Tyler Muhly's **CLUS road simulation example** will be again used within this documentation.

In the context of this package, resource development scenarios are represented by three components:

1. a spatial, gridded cost layer covering the entire landscape, wherein each cell value represents the cost of new road developement through it.
2. a spatial, representation of the existing road network within the landscape, and
3. a set of locations, refered to as "**landings**", where new resource development is to take place.

The nature of how the landings are defined in the resource development scenario dictates whether the new roads development simulation will be a one-time or multi-temporal simulation.

Given an input resource development scenario, the `projectRoads` method will simulate new roads between the existing road network and the landings using one of three user-specific simulation methods (not to be confused with R language methods):

1. the **snapping** simulation method,
2. the **least-cost path (LCP)** simulation method, or
3. the **minimum spanning tree (MST)** simulation method.


# Resource development scenario

### 1. Cost layer

The cost of new roads development within a landscape is specified to the `projectRoads` method in the form of a single, numeric [`RasterLayer`](https://www.rdocumentation.org/packages/raster/versions/2.9-5/topics/Raster-class) {[`raster`](https://cran.r-project.org/web/packages/raster/index.html)} object.  Each cell of this raster represents the cost of new road developement (the definition of which is up to the user) through it.  Following this, the total cost of the development of a new road segment is considered to be the sum of the cost raster cells that it traverses.

* The traversal cost of existing road segments is assumed to be zero.
* The extent of the landscape of interest is assumed to match the extent of the specified cost raster layer.
* The coordinate reference system (CRS) of this raster does not need to be specified, so long as all other resource development scenario input data-sets match.  
* While the cell size (or resolution) of this cost raster is up the user, it is important to remember that the a unit area of road is assumed to take up an entire cell.

The simple cost raster used in the CLUS roads simulation example is included within the `CLUSexample` data-set of this package.

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=5,fig.height=4.85)
```

```{r, fig.show='hold'}
#library(roads)
devtools::load_all(".")
library(raster)
library(colorspace)
## colours for displaying cost raster (optional; could just use default)
rastColours <- c('grey50',diverge_hcl(20,h=c(130,43),c=100,l=c(70,90)))
## get the cost raster from the CLUSexmaple data-set and plot
costRaster <- roads::CLUSexample$cost
plot(costRaster, col = rastColours, main = 'Cost')
```

Notice that the top row of the raster has a cost of zero.  This row is where, in this scenario, an existing road traverses the landscape.

### 2. Existing road network layer

Since, in the context of this package, new roads simulation involves connecting landings to an existing roads network, the state of the existing roads network must be specified to the `projectRoads` method.  This specification is made in the form of a sf object with geometry type "LINES", a SpatialLines object, or a RasterLayer.

The existing roads network used in the CLUS roads simulation example is included within the `CLUSexample` data-set of this package.  Alternatively, it could be extracted from cost raster layer by isolating cells where the cost is zero.

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=7,fig.height=3.93)
```

```{r, fig.show='hold'}
## existing roads network raster from the CLUSexample data-set
existingRoadsRast <- roads::CLUSexample$roads
roadsLine <- roads::CLUSexample$roads %>% rasterToLineSegments()
## alternatively, existing roads network raster from the cost raster layer, where cells are zero
alt_existingRoadsRast <- costRaster==0
## they are the same?
raster::compareRaster(existingRoadsRast,alt_existingRoadsRast,
                      values=TRUE,extent=TRUE,res=TRUE,crs=TRUE)
## the existing roads raster matches the cost raster in terms of extent, resolution, and CRS?
raster::compareRaster(existingRoadsRast,costRaster,
                      values=FALSE,extent=TRUE,res=TRUE,crs=TRUE)
## plot cost ratser (left) and existing roads raster (right)
par(mfrow=c(1,2))
plot(costRaster,col=rastColours,main='Cost')
plot(existingRoadsRast,col=c('grey80','grey50'),main='Current roads',legend=FALSE)
text(x=2.5,y=4.5,labels='existing road',cex=1.3)
text(x=2.5,y=2,labels='not currently\nroad',cex=1.3)
```

## 3. Landings layer(s)

Landings, or resource development locations, that are to be connected to the existing road network can be specified to the `projectRoads` method in a variety of forms.  These include specification as:

* A logical [`RasterLayer`](https://www.rdocumentation.org/packages/raster/versions/2.9-5/topics/Raster-class) {[`raster`](https://cran.r-project.org/web/packages/raster/index.html)} object with cell values of `TRUE` for landings,
* A [`SpatialPoints`](https://www.rdocumentation.org/packages/sp/versions/1.3-1/topics/SpatialPoints) {[`sp`](https://cran.r-project.org/web/packages/sp/index.html)} or [`SpatialPointsDataFrame`](https://www.rdocumentation.org/packages/sp/versions/1.3-1/topics/SpatialPointsDataFrame-class) {[`sp`](https://cran.r-project.org/web/packages/sp/index.html)} object with points representing landings,
* A matrix with at least two columns, with columns 1 and 2 representing x any y coordinates of landing locations respectively, 
* A [`SpatialPolygons`](https://www.rdocumentation.org/packages/sp/versions/1.3-1/topics/SpatialPolygons-class) {[`sp`](https://cran.r-project.org/web/packages/sp/index.html)} or [`SpatialPolygonsDataFrame`](https://www.rdocumentation.org/packages/sp/versions/1.3-1/topics/SpatialPolygonsDataFrame-class) {[`sp`](https://cran.r-project.org/web/packages/sp/index.html)} object with polygons representing landings, 
* A sf object with geometry type "LINES" or "POLYGONs" or,
* A [`RasterStack`](https://www.rdocumentation.org/packages/raster/versions/2.9-5/topics/Raster-class) {[`raster`](https://cran.r-project.org/web/packages/raster/index.html)} or [`RasterBrick`](https://www.rdocumentation.org/packages/raster/versions/2.9-5/topics/Raster-class) {[`raster`](https://cran.r-project.org/web/packages/raster/index.html)} object with each raster layer in the stack or brick representing a new time step for which new roads are to be simulated for (i.e. layer 1 represents landings at time t=1, layer 2 represents landings at time t=2, and so on).  Within each raster layer, cell values of `TRUE` represent new landings for that time-step.

The set of landings used in the CLUS roads simulation example is included within the CLUSexample data-set of this package.

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=5,fig.height=4.85)
```

```{r, fig.show='hold'}
## landings as spatial points
landings <- roads::CLUSexample$landings
## plot landings on top of cost raster
plot(costRaster,col=rastColours,main='Cost with landings')
plot(landings,add=TRUE,pch=19)
text(x=2.5,y=4.5,labels='existing road',cex=1.3)
points(x=5.6,y=4.5,pch=19,xpd=TRUE)
text(x=5.8,y=4.5,labels='landing',adj=c(0,0.4),xpd=TRUE)
```

# Simulation of new roads development

## Simulation methods

### 1. Snapping approach/method

This approach simply 'snaps' a landing (i.e., cutblock location) to the nearest existing road segment. Since the snapping is done for each landing it is also called an independent path method. 


```{r, fig.show='hold'}
## project new roads using the 'snap' approach
projRoads_snap <- roads::projectRoadsNew(landings,
                                         costRaster,
                                         roadsLine,
                                         roadMethod='snap')
## plot the cost raster, landings, and roads segments to the landings
plot(costRaster, col = rastColours, main = '\'Snapped\' roads')
points(landings, pch = 19)  ## landings points
plot(projRoads_snap$roads, add = TRUE) ## roads
## update legend
points(x = 5.5, y = 4.8, pch = 19, xpd = TRUE)
text(x=5.7,y=4.8,labels='landing',adj=c(0,0.4),xpd=TRUE)
lines(x=c(5.3,5.7),y=c(4.2,4.2),lwd=2,xpd=TRUE)
text(x=5.75,y=4.2,labels='roads',adj=c(0,0.3),xpd=TRUE)
```

Using this approach, a few issues would arise: 

1. parallel roads are not realistic since there is no branching and this ultimately leads to increases in the amount of roads; 

2. costs are not included (i.e., slope and barriers like large waterbodies).

This means this approach, while simple to implement, would over estimate the amount of simulated roads.

### 2. Least Cost Paths (LCP) approach /method

This approach builds upon the snapping approach by assuming a 'cost directed' path (i.e., "as the wolf runs") for each landing to the existing road network. This includes the cost of barriers and impedance attributes. However, this approach requires a cost surface that needs to be estimated, then updated within the simulations and thus, requires greater computational time relative to the snapping approach. This data.table is then used to build a mathematical [graph](https://en.wikipedia.org/wiki/Graph_theory) using [igraph](http://igraph.org/r/).

Once the graph is built, it is simple to get the least cost path between any two verticies. Using [Dijkstra’s algorithm](https://en.wikipedia.org/wiki/Dijkstra%27s_algorithm), the [shortest_paths](http://igraph.org/r/doc/distances.html) function is called in [igraph](http://igraph.org/r/doc/#R). Further, the graph can be easily updated to simulate the change in the cost surface from dynamically developing roads during the simulation. Note, that this dynamic behaviour would not be possible in a pre-solve of the road network. To implement a pre-solve approach the [igraph](http://igraph.org/r/doc/#R) function [all_shortest_paths](http://igraph.org/r/doc/distances.html) can be used to calculate all shortest paths between pairs of vertices; however, this is not included in the comparison.

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=7,fig.height=3.93)
```


```{r, fig.show='hold'}
## project new roads using the 'LCP' approach
projRoads_lcp <- roads::projectRoadsNew(landings, 
                                        costRaster, 
                                        roadsLine, 
                                        roadMethod = 'lcp')

## plot the cost raster and overlay it with new roads
plot(costRaster,col=rastColours,main='Cost with new roads')
plot(projRoads_lcp$roads, add = TRUE)
points(landings,pch=19,col='red')  ## landings points
## legend
lines(c(-0.5, 0), c(-1.45, -1.45), xpd = TRUE)
text(x=0.1,y=-1.45,labels='new road',adj=c(0,0.4),xpd=TRUE)
points(x=2.5,y=-1.45,pch=19,col='red',xpd=TRUE)
text(x=2.7,y=-1.45,labels='landing',adj=c(0,0.4),xpd=TRUE)
```

The main disadvantage of this approach is the view of independence of road development. The least cost path may produce parallel or redundant roads given a path is made for each target to the corresponding closest point. This may mimic road development given road tenures are able to all licensees the right to limit other industrial users from using their road (i.e., gated roads); thereby forcing the other industrial user to consider building a nearly paralell road. In some cases there will be branching, where two roads connecting two landings to an existing road network will use the same least cost path; however, this will be conditional on the spatial configuration of the local cost surface and the existing road network. Thus, the amount of road being developed from the LCP is dependant on the local cost surface and may be either higher or lower than the corresponding snap approach.

### 3. Minimum Spanning Tree (MST) with Least Cost Paths (LCP) approach/method

The MST approach builds upon the LCP approach by determining if landings should be connected to one another before being connected to the existing road network. In the MST approach, LCPs are estimated both between the landings and between landings and the existing road network. These distances are then used as nodes for solving a minimum spanning tree (MST). The sequence of verticies from the LCPs are then constructed following the solution to the MST. This approach allows for greater branching in simulated roads which would reduce the road building cost relative to the LCP approach. However, the computational time would increase over the LCP approach given the extra steps of getting LCP distances and then solving the MST.

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=7,fig.height=3.93)
```

```{r, fig.show='hold'}
## project new roads using the 'MST' approach
projRoads_mst <- roads::projectRoadsNew(landings, 
                                        costRaster,
                                        roadsLine, 
                                        roadMethod = 'mst')

## plot the cost raster and overlay it with new roads
plot(costRaster, col = rastColours, main = 'Cost with new roads')
plot(projRoads_mst$roads, col = 'grey30', add = TRUE, legend = FALSE)
points(landings, pch = 19, col = 'red')  ## landings points
## legend
lines(c(-0.5, 0), c(-1.45, -1.45), xpd = TRUE)
text(x=0.1,y=-1.45,labels='new road',adj=c(0,0.4),xpd=TRUE)
points(x=2.5,y=-1.45,pch=19,col='red',xpd=TRUE)
text(x=2.7,y=-1.45,labels='landing',adj=c(0,0.4),xpd=TRUE)
```

The MST approach will produce the least amount of roads (relative to the other approaches), given targets are allowed to connect to other target locations. This approach simulates a realistic view of road branching relative to the other two approaches. However, given the need to get LCP distances, solve a MST and then construct the LCPs, it will likely be the most costly in terms of computation time.

## One-time versus multi-temporal simulation

Roads development simulation can be performed either for a single time step (one-time) or for multiple time steps.  This section will use a demonstration scenario data-set that is included with the roads package.

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=7,fig.height=6.5)
```

```{r, fig.show='hold'}
## colours for displaying cost raster (optional; could just use default)
library(colorRamps)
rastColours2 <- c('grey50', colorRamps::matlab.like(300)[25:275])
## scenario 
scen <- roads::demoScen[[1]]
## landing sets 1 to 4 of this scenario 
land.pnts <- scen$landings.points[scen$landings.points$set%in%c(1:4),]
## plot the cost raster and landings
par(mar=par('mar')/2)
plot(scen$cost.rast, col = rastColours2, main = 'Cost and landings (by set)')
plot(land.pnts, add = TRUE, pch = 21, cex = 2, bg = 'white')
text(land.pnts@coords, labels = land.pnts$set, cex = 0.6, adj = c(0.5, 0.3),
     xpd = TRUE)
```

### One-time simulation

If landings, costs, and roads are all specified to `projectRoads`, then a one-time road simulation will be performed that returns a list object holding the projected roads and related information. This can be repeated multiple time for different road building scenarios.

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=7,fig.height=6.86)
```

```{r, fig.show='hold'}


## project roads for landing sets 1 to 4, with independent one-time simulations
oneTime_sim <- list() ## empty raster stack will be updated with sim results rasters
for (i in 1:4){
  oneTime_sim <- c(oneTime_sim,
                       roads::projectRoadsNew(land.pnts[land.pnts$set==i,],
                                              scen$cost.rast,
                                              scen$cost.rast==0,
                                              roadMethod='mst')$roads)
}
#oneTime_sim[oneTime_sim==0 | scen$cost.rast==0] <- NA ## set non-road to NA for display purposes

## plot
par(mfrow = c(2, 2))
par(mar = par('mar')/2)
for (i in 1:4){
  plot(scen$cost.rast, col = rastColours2, 
       main = paste0('Landings set ', i),
       legend = FALSE)
  plot(oneTime_sim[[i]], add = TRUE)#,
       # col = 'grey30', 
       # add = TRUE, legend = FALSE)
  plot(land.pnts[land.pnts$set == i, ], add = TRUE,
       pch = 21, cex = 1.5, bg = 'white')
}

```

While the results of these one-time simulations may be fine when looking at each landings scenario/set idependently (e.g. each representing a possible scenario for time t=1), they are likely not appropriate for cases where all landings sets follow a temporal development sequence (e.g. set 1 is development at time t=1, set 2 is development at time t=2, and so on). Independent one-time simulations do not take into account the fact that, for a given time step, existing roads for a given simulation should be the union of existing roads at time t=0 and all roads simulation results leading up to the current step.  For example, existing roads input into simulation at time t=2 (landings set 2), would be the union of existing roads at time t=0 and projected roads at time t=1 (landings set 1).

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=5,fig.height=4.85)
```

```{r, fig.show='hold'}
## raster representing the union of completely independent simulations for multiple sets
oneTime_sim <- raster::stack(oneTime_sim)
independent <- any(oneTime_sim == 1)
independent[!independent] <- NA ## set non-road to NA for display purposes
## plot 
plot(scen$cost.rast,col=rastColours2,main='Union of independent sim results',legend=FALSE)
plot(independent,col='grey30',add=TRUE,legend=FALSE)
plot(land.pnts,add=TRUE,pch=21,cex=1.5,bg='white')
```

### Multi-temporal simulation

Multi-temporal (multiple time steps) roads projections can be performed by `projectRoads`, by providing the list produced by a previous simulation run (sim list) and the landings for the current time period. The function uses the sim list as a starting point and does not need to recompute the graph used to determine the least cost path. This can be implemented in a loop.

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=7,fig.height=9.97)
```


```{r, fig.show='hold'}
## continuing on with demo scenario 1
## landing sets 1 to 4 of this scenario as a raster stack
land.stack <- scen$landings.stack[[1:4]]
## pass the landings RasterStack to the projectRoads method for multi-temporal simulation
# initialize sim list with first landings set
multiTime_sim <- list(projectRoadsNew(land.stack[[1]], scen$cost.rast, 
                                               scen$road.line))

# iterate over landings sets using the sim list from the previous run as input
for (i in 2:nlayers(land.stack)) {
  multiTime_sim <- c(
    multiTime_sim,
    list(projectRoadsNew(sim =  multiTime_sim[[i-1]], landings = land.stack[[i]]))
  ) 
}


par(mfrow = c(3, 2))
par(mar = par('mar')/2)
  plot(scen$cost.rast, col = rastColours2, main = 'Roads at time t=0', 
       legend = FALSE)
  plot(scen$road.line, col = 'grey30', add = TRUE, legend = FALSE)
  
for (i in 1:length(multiTime_sim)){
  plot(scen$cost.rast, col = rastColours2, 
       main = paste0('Roads at time t=', i), legend = FALSE)
  plot(multiTime_sim[[i]]$roads, col = 'grey30', add = TRUE, legend = FALSE)
  plot(land.pnts[land.pnts$set == i, ], add = TRUE, pch = 21, 
         cex = 1.5, bg = 'white')
  if (i >= 2){
    plot(land.pnts[land.pnts$set < i, ], add = TRUE, pch = 1, cex = 1.5)
    plot(land.pnts[land.pnts$set == i, ], add = TRUE, pch = 21, 
         cex = 1.5, bg = 'white')
  }
}
```






