---
title: "roads Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{roads Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup1, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# ignore spherical coordinates 
s2_in <- sf::sf_use_s2(FALSE)
```

# Introduction

This vignette provides a tutorial on the use of the `roads` package for the spatial simulation of future roads development under a given resource development scenario.  This tutorial, which borrows heavily from a demonstration written by Kyle Lochhead and Tyler Muhly in 2018, will focus primarily on the `projectRoads` function of the `roads` package, as it is responsible for performing the simulation. Example data sets used below are included in the package as `CLUSexample` and `demoScen`. 

In the context of this package, resource development scenarios are represented by three components:

1. **Weights Raster and Weight Function:** the weights raster is a spatial, gridded weight layer covering the entire landscape. The weight function takes values from adjacent weights raster cells as inputs and returns the cost of developing a new road between them.
2. **Existing Road Network:** a spatial, representation of the existing road network within the landscape.
3. **Landings:** a set of locations where new resource development is to take place.

Given an input resource development scenario, `projectRoads` will simulate new roads between the existing road network and the landings using one of four algorithms:

1. **Snapping** 
2. **Least-cost path (LCP)** 
3. **Iterative least-cost path (ILCP)** 
4. **Minimum spanning tree (MST)** 

The output of `projectRoads` is a list of simulation results referred to as a "sim list". The list contains five elements:

1. **roads:** the projected road network, including new and input roads.
2. **weightRaster:** the weights raster, updated to reflect the new roads that were added.
3. **roadMethod** the road simulation method used.
4. **landings** the landings used in the simulation.
5. **g** the graph that describes the cost of paths between each cell in the weights raster. This is updated based on the new roads so that vertices connected by new roads now have a weight of 0. This can be used to avoid recomputing the graph in a simulation with multiple time steps. 

### Setup

```{r setup, results='hide', warning=FALSE, message=FALSE}
library(terra)
library(dplyr)
library(sf)
library(roads)

## colours for displaying weight raster 
if(requireNamespace("viridis", quietly = TRUE)){
  # Use colour blind friendly palette if available
  rastColours <- c('grey50', viridis::viridis(20))
} else {
  rastColours <- c('grey50', terrain.colors(20))
}

# terra objects need to be wrapped to be saved, this unwraps them
CLUSexample <- prepExData(CLUSexample)
```


# Resource development scenario

### 1. Weights Raster and Weight Function

The cost of new roads development within the landscape is determined by applying the `weightFunction` to the `weightRaster`. The `weightRaster` must be provided as a single, numeric `SpatRaster` or `RasterLayer` object.  The cell values in the raster for two adjacent cells are provided to the `weightFunction` to calculate the weight for the edge connecting the two cells in the graph which should represent the cost of building a road connecting the two cells. Two ways of calculating the cost are supported by the package with the option for users to develop their own. The simplest method is for the `weightRaster` values to represent the cost of building a road across a cell and the for the `weightFunction` to simply take the mean of the values in adjacent cells. The other option included in the package is to use the `slopePenaltyFn`. This function takes a modified DEM raster as input and determines costs by applying a base cost and a grade penalty which is multiplied by the % grade between adjacent cells. The DEM can be modified by adding barriers to the raster, either as negative values which represent high costs (e.g. a stream crossing) or NAs for inaccessible areas (e.g. a cliff or a lake). See `?slopePenaltyFn` for more details.   

The following points apply to all `weightRaster`s:  
* Weights of existing road segments are assumed to be zero. If existing roads are not 0 in the `weightRaster` they can be "burned in" by setting `roadsInWeight = FALSE`.
* The extent of the landscape of interest is assumed to match the extent of the specified `weightRaster`.
* The coordinate reference system (CRS) of this raster must match all other input data-sets.  
* While the cell size (or resolution) of the `weightRaster` is up to the user, it is important to remember that a unit area of road is assumed to take up an entire cell.
* Areas that are NA are not included in the graph and are assumed to be inaccessible to roads. If a landing is surrounded by NAs then it cannot be connected to the existing road network and an error is produced.

The weight raster for this basic example is a cost surface and the `weightFunction` used is the average. 
```{r}
weightRaster <- CLUSexample$cost
```

### 2. Existing road network layer

The state of the existing roads network must be specified to the `projectRoads` method. This specification is made in the form of an `sf` object with geometry type "LINES", a `SpatialLines` object, a `RasterLayer`, or a `SpatRaster`. If the input roads are a raster the resulting roads will be returned as a raster by default but if `roadsOut = "sf"` is set then a geometry collection of lines and points will be returned with points representing the input roads.

The roads network included in the `CLUSexample` data-set is a raster but we use a line to make plotting easier.  

```{r, fig.show='hold'}
## existing roads network
roadsLine <- sf::st_sfc(geometry = sf::st_linestring(
  matrix(c(0.5, 4.5, 4.5, 4.5),
         ncol = 2, byrow = T) 
)) %>%
  sf::st_as_sf(crs = sf::st_crs(weightRaster))
```

### 3. Landings layer(s)

Landings, or resource development locations, that are to be connected to the existing road network can be specified to the `projectRoads` method in a variety of forms.  These include specification as:

* A `sf` object with geometry type "POINTS" or "POLYGONs"
* A logical `SpatRaster` or `RasterLayer` object with cell values of `TRUE` for landings,
* A `SpatialPoints` or `SpatialPointsDataFrame` object with points representing landings,
* A matrix with at least two columns, with columns 1 and 2 representing x any y coordinates of landing locations respectively, 
* A `SpatialPolygons`or `SpatialPolygonsDataFrame` object with polygons representing landings, 


If the landings are polygons then the centroid is used as the destination for new roads. For more control or to have more than one landing per polygon see Multiple landings per harvest block below.

```{r, fig.show='hold', fig.width=5, fig.height=4.85}
## landings as spatial points
landings <- roads::CLUSexample$landings

## plot example scenario
plot(weightRaster, col = rastColours, main = 'Example Scenario')
plot(roadsLine, add = TRUE)
plot(landings, add = TRUE, pch = 19)
points(x=5.6,y=4.5,pch=19,xpd=TRUE)
text(x=5.8,y=4.5,labels='landing',adj=c(0,0.4),xpd=TRUE)
lines(x=c(5.3,5.6),y=c(4.2,4.2),lwd=2,xpd=TRUE)
text(x=5.75,y=4.2,labels='roads',adj=c(0,0.3),xpd=TRUE)

```

Notice that the top row of the raster has a cost of zero, this is where an existing road traverses the landscape.

### Output format
`projectRoads` accepts a wide range of classes of spatial objects as input but all results are returned as `sf` for vectors and `SpatRaster` for rasters. 

# Simulation of new roads development

## Simulation methods

### 1. Snapping

This approach simply 'snaps' a landing to the nearest existing road segment. Since the snapping is done for each landing it is also called an independent path method. 

```{r, fig.show='hold', fig.width=5, fig.height=4.85}
## project new roads using the 'snap' approach
projRoads_snap <- roads::projectRoads(landings, weightRaster, roadsLine,
                                      roadMethod = 'snap')

## plot the weight raster, landings, and roads segments to the landings
plot(weightRaster, col = rastColours, main = "'Snapped' roads")
points(landings, pch = 19, col = 'red')  
plot(projRoads_snap$roads, add = TRUE) 

## update legend
points(x = 5.5, y = 4.8, pch = 19, xpd = TRUE, col = 'red')
text(x = 5.7, y = 4.8, labels = 'landing', adj = c(0, 0.4), xpd = TRUE)
lines(x = c(5.3, 5.6), y = c(4.2, 4.2), lwd = 2, xpd = TRUE)
text(x = 5.75, y = 4.2, labels = 'roads', adj = c(0, 0.3), xpd = TRUE)
```

Using this approach, a few issues would arise: 

1. parallel roads are not realistic since there is no branching and this leads to increased numbers of roads; 

2. costs are not included (i.e., slope and barriers like large water bodies).

This means this approach, while simple to implement, would over estimate the amount of simulated roads.

### 2. Least Cost Paths (LCP)

This approach builds upon the snapping approach by assuming a 'cost directed' path (i.e., "as the wolf runs") for each landing to the existing road network.  This approach requires that a cost surface be provided and used to build a mathematical [graph](https://en.wikipedia.org/wiki/Graph_theory) using [`igraph`](https://igraph.org/r/) and takes considerably longer to compute.

Once the graph is built, the least cost path between each landing and the nearest existing road is determined using [Dijkstra’s algorithm](https://en.wikipedia.org/wiki/Dijkstra%27s_algorithm), which is implemented in the [shortest_paths](https://igraph.org/r/doc/distances.html) function in `igraph`. Then the graph is updated so that the graph returned in the result reflects the new roads built.

```{r, fig.show='hold', fig.width=5, fig.height=4.85}
## project new roads using the 'LCP' approach
projRoads_lcp <- roads::projectRoads(landings, 
                                        weightRaster, 
                                        roadsLine, 
                                        roadMethod = 'lcp')

## plot the weight raster and overlay it with new roads
plot(weightRaster, col = rastColours, main = "'LCP' roads")
plot(projRoads_lcp$roads, add = TRUE)
points(landings, pch = 19, col = 'red')  ## landings points
## legend
points(x = 5.5, y = 4.8, pch = 19, xpd = TRUE, col = 'red')
text(x = 5.7, y = 4.8, labels = 'landing', adj = c(0, 0.4), xpd = TRUE)
lines(x = c(5.3, 5.6), y = c(4.2, 4.2), lwd = 2, xpd = TRUE)
text(x = 5.75, y = 4.2, labels = 'roads', adj = c(0, 0.3), xpd = TRUE)
```

The main disadvantage of this approach is that roads are developed independently. The least cost path may produce parallel or redundant roads since a path is made for each target to the corresponding closest point. This may mimic road development since road tenures give licensees the right to limit other industrial users from using their road (i.e., gated roads); thereby forcing the other industrial users to consider building a nearly parallel road. In some cases there will be branching, where two roads connecting two landings to an existing road network will use the same least cost path; however, this will be conditional on the spatial configuration of the local cost surface and the existing road network. Thus, the amount of road being developed from the LCP is dependent on the local cost surface and may be either higher or lower than the corresponding snap approach.

### 3. Iterative Least Cost Paths (ILCP)
This approach reduces the number of parallel roads seen in the LCP method because a road to each landing is built sequentially and the cost in the graph is updated after each so that roads built earlier can be used to access other landings. The order is determined by the `ordering` argument to `projectRoads` and by default builds roads to the closest landings first. To prevent this and build roads in the order that the landings are supplied use `ordering = none`. 

```{r, fig.show='hold', fig.width=5, fig.height=4.85}
## project new roads using the 'ILCP' approach
projRoads_ilcp <- roads::projectRoads(landings, 
                                        weightRaster, 
                                        roadsLine, 
                                        roadMethod = 'ilcp')

## plot the weight raster and overlay it with new roads
plot(weightRaster, col = rastColours, main = "'ILCP' roads")
plot(projRoads_ilcp$roads, add = TRUE)
points(landings, pch = 19, col = 'red')  ## landings points
## legend
points(x = 5.5, y = 4.8, pch = 19, xpd = TRUE, col = 'red')
text(x = 5.7, y = 4.8, labels = 'landing', adj = c(0, 0.4), xpd = TRUE)
lines(x = c(5.3, 5.6), y = c(4.2, 4.2), lwd = 2, xpd = TRUE)
text(x = 5.75, y = 4.2, labels = 'roads', adj = c(0, 0.3), xpd = TRUE)
```

The ILCP approach produces more of a network rather than many parallel roads. However, it is sensitive to the ordering of the landings. Below we reverse the order of the landings but continue using the default ordering of closest first. The two closest landings are tied for distance to the road and the tie is broken by the order they are supplied in so switching that produces a different road network.   

```{r, fig.show='hold', fig.width=5, fig.height=4.85}
## project new roads using the 'ILCP' approach
projRoads_ilcp2 <- roads::projectRoads(st_coordinates(landings)[4:1,], 
                                        weightRaster, 
                                        roadsLine, 
                                        roadMethod = 'ilcp')

## plot the weight raster and overlay it with new roads
plot(weightRaster, col = rastColours, main = "'ILCP' roads")
plot(projRoads_ilcp2$roads, add = TRUE)
points(landings, pch = 19, col = 'red')  ## landings points
## legend
points(x = 5.5, y = 4.8, pch = 19, xpd = TRUE, col = 'red')
text(x = 5.7, y = 4.8, labels = 'landing', adj = c(0, 0.4), xpd = TRUE)
lines(x = c(5.3, 5.6), y = c(4.2, 4.2), lwd = 2, xpd = TRUE)
text(x = 5.75, y = 4.2, labels = 'roads', adj = c(0, 0.3), xpd = TRUE)
```

### 4. Minimum Spanning Tree (MST) with Least Cost Paths (LCP)

The MST approach builds upon the LCP approach by determining if landings should be connected to one another before being connected to the existing road network. In the MST approach, LCPs are estimated both between the landings and between landings and the existing road network. These distances are then used as nodes for solving a minimum spanning tree. The sequence of vertices from the LCPs are then constructed following the solution to the MST. 

```{r, fig.show='hold',  fig.width=5, fig.height=4.85}
## project new roads using the 'MST' approach
projRoads_mst <- roads::projectRoads(landings, 
                                        weightRaster,
                                        roadsLine, 
                                        roadMethod = 'mst')

## plot the weight raster and overlay it with new roads
plot(weightRaster, col = rastColours, main = "'MST' roads")
plot(projRoads_mst$roads, add = TRUE)
points(landings, pch = 19, col = 'red')  ## landings points
## legend
points(x = 5.5, y = 4.8, pch = 19, xpd = TRUE, col = 'red')
text(x = 5.7, y = 4.8, labels = 'landing', adj = c(0, 0.4), xpd = TRUE)
lines(x = c(5.3, 5.6), y = c(4.2, 4.2), lwd = 2, xpd = TRUE)
text(x = 5.75, y = 4.2, labels = 'roads', adj = c(0, 0.3), xpd = TRUE)
```

The MST approach will produce the least amount of roads (relative to the other approaches), given landings are allowed to connect to other landings. This approach simulates a realistic view of road branching relative to the other approaches. However, given the need to get LCP distances, solve a MST and then construct the LCPs, it will likely be the most costly in terms of computation time.

## One-time versus multi-temporal simulation

Roads development simulation can be performed either for a single time step (one-time) or for multiple time steps.  This section will use a demonstration scenario `demoScen` data-set that is included with the roads package. There are four different sets of landings. 

```{r, fig.show='hold', fig.width=7,fig.height=6.5}
## colours for displaying weight raster
if(requireNamespace("viridis", quietly = TRUE)){
  # Use colour blind friendly palette if available
  rastColours2 <- c('grey50', viridis::viridis(30))
} else {
  rastColours2 <- c('grey50', terrain.colors(30))
}

## scenario 
demoScen <- prepExData(demoScen)
scen <- demoScen[[1]]
## landing sets 1 to 4 of this scenario 
land.pnts <- scen$landings.points[scen$landings.points$set %in% c(1:4),]
## plot the weight raster and landings
par(mar=par('mar')/2)
plot(scen$cost.rast, col = rastColours2, main = 'Cost and landings (by set)')
plot(land.pnts %>% st_geometry(), add = TRUE, pch = 21, cex = 2, bg = 'white')
text(st_coordinates(land.pnts), labels = land.pnts$set, cex = 0.6, adj = c(0.5, 0.3),
     xpd = TRUE)
```

### One-time simulation

If landings, costs, and roads are all specified to `projectRoads`, then a one-time road simulation will be performed that returns a list object holding the projected roads and related information. This can be repeated multiple times for different road building scenarios but each simulation will be independent of previous simulations. This would be appropriate if each landing set represented alternate scenarios for development.

```{r, fig.show='hold', fig.width=7,fig.height=6.86}
## project roads for landing sets 1 to 4, with independent one-time simulations
oneTime_sim <- list() ## empty list 
for (i in 1:4){
  oneTime_sim <- c(oneTime_sim,
                       roads::projectRoads(land.pnts[land.pnts$set==i,],
                                              scen$cost.rast,
                                              scen$cost.rast==0,
                                              roadMethod='mst')$roads)
}

## plot
oldpar <- par(mfrow = c(2, 2), mar = par('mar')/2)
for (i in 1:4){
  oneTime_sim[[i]][!oneTime_sim[[i]]] <- NA 
  plot(scen$cost.rast, col = rastColours2, 
       main = paste0('Landings set ', i),
       legend = FALSE)
  plot(oneTime_sim[[i]], add = TRUE, col = "grey50", legend = FALSE)
  plot(st_geometry(land.pnts[land.pnts$set == i, ]), add = TRUE,
       pch = 21, cex = 1.5, bg = 'white')
}

```

While the results of these one-time simulations may be fine when looking at each landings scenario/set independently (e.g. each representing a possible scenario for time t=1), they are likely not appropriate for cases where all landings sets follow a temporal development sequence (e.g. set 1 is development at time t=1, set 2 is development at time t=2, and so on). Independent one-time simulations do not take into account the fact that, for a given time step, existing roads for a given simulation should be the union of existing roads at time t=0 and all roads simulation results leading up to the current step.  For example, existing roads input into simulation at time t=2 (landings set 2), would be the union of existing roads at time t=0 and projected roads at time t=1 (landings set 1).

```{r, fig.show='hold', fig.width=5,fig.height=4.85}
## raster representing the union of completely independent simulations for multiple sets
oneTime_sim <- rast(oneTime_sim)
independent <- any(oneTime_sim, na.rm = TRUE)
## set non-road to NA for display purposes
independent[!independent] <- NA

## plot 
plot(scen$cost.rast, col = rastColours2,
     main = 'Union of independent sim results',
     legend = FALSE)

plot(independent, col = 'grey30', add = TRUE, legend = FALSE)

plot(st_geometry(land.pnts), add = TRUE, pch = 21, cex = 1.5, bg = 'white')
```

### Multi-temporal simulation

Multi-temporal (multiple time steps) roads projections can be performed by `projectRoads`, by providing the list produced by a previous simulation run (sim list) to the `sim` argument and the landings for the current time period. The function uses the sim list as a starting point and does not need to recompute the graph used to determine the least cost path. This can be implemented in a loop.

```{r, fig.show='hold', fig.width=7, fig.height=9.97}
## continuing on with demo scenario 1
## landing sets 1 to 4 of this scenario as a raster stack
land.stack <- scen$landings.stack[[1:4]]

# initialize sim list with first landings set
multiTime_sim <- list(projectRoads(land.stack[[1]], scen$cost.rast, 
                                               scen$road.line))

# iterate over landings sets using the sim list from the previous run as input
for (i in 2:nlyr(land.stack)) {
  multiTime_sim <- c(
    multiTime_sim,
    list(projectRoads(sim =  multiTime_sim[[i-1]], landings = land.stack[[i]]))
  ) 
}

par(mfrow = c(3, 2))
par(mar = par('mar')/2)
  plot(scen$cost.rast, col = rastColours2, main = 'Roads at time t = 0', 
       legend = FALSE)
  plot(scen$road.line, col = 'grey30', add = TRUE, legend = FALSE)
  
for (i in 1:length(multiTime_sim)){
  plot(multiTime_sim[[i]]$weightRaster, col = rastColours2, 
       main = paste0('Roads at time t = ', i), legend = FALSE)
  plot(multiTime_sim[[i]]$roads, col = 'grey30', add = TRUE, legend = FALSE)
  plot(st_geometry(land.pnts[land.pnts$set == i, ]), add = TRUE, pch = 21, 
         cex = 1.5, bg = 'white')
  if (i >= 2){
    plot(st_geometry(land.pnts[land.pnts$set < i, ]), add = TRUE, pch = 1, cex = 1.5)
    plot(st_geometry(land.pnts[land.pnts$set == i, ]), add = TRUE, pch = 21, 
         cex = 1.5, bg = 'white')
  }
}
```


## Multiple landings per harvest block

Often harvest information is available as polygons showing the cutover area but the point locations of landings are not known. The roads package includes the `getLandingsFromTarget` function to address these situations. By default `getLandingsFromTarget` will use the centroid of a polygon as the landing but you can also specify a `sampleType` of "random" or "regular" and a `landingDens` in landings per unit area to generate multiple landing points with in the harvest block.

```{r fig.width=6,fig.height=5}
harvPoly <- demoScen[[1]]$landings.poly

outCent <- getLandingsFromTarget(harvPoly)
plot(sf::st_geometry(harvPoly))
plot(outCent, col = "red", add = TRUE)

# Get random sample with density 0.02 pts per unit area
outRand <- getLandingsFromTarget(harvPoly, 0.02, sampleType = "random")
prRand <- projectRoads(outRand, scen$cost.rast, scen$road.line)

plot(scen$cost.rast, main = "Random Landings in Harvest Blocks",
     col = rastColours2)
plot(harvPoly, add = TRUE)
plot(prRand$roads, add = TRUE,  col = "grey50")
plot(outRand, col = "red", add = TRUE)

# Get regular sample with density 0.02 pts per unit area
outReg <- getLandingsFromTarget(harvPoly, 0.02, sampleType = "regular")
prReg <- projectRoads(outReg, scen$cost.rast,scen$road.line)


plot(scen$cost.rast, main = "Regular Landings in Harvest Blocks",
     col = rastColours2)
plot(harvPoly, add = TRUE)
plot(prReg$roads, add = TRUE, col = "grey50")
plot(outReg, col = "red", add = TRUE)


# clean up 
par(oldpar)
```

The regular sampling method may be the most realistic since it ensures that landings are spaced apart from each other.

# Note
This vignette is partially copied from Kyle Lochhead & Tyler Muhly's [2018 CLUS example](https://github.com/bcgov/castor/blob/main/reports/roads/draft-CLUS-roads.md)

```{r, include=FALSE}
# ignore shperical coordinates 
sf::sf_use_s2(s2_in)
```

